---
title: "Group 6 Data Analysis"
format: 
  html:
    code-fold: true
editor: visual
---

# 1 Introduction

Online movie rating platforms such as IMDB provide valuable insights into audience opinions by collecting user ratings and reviews, which many viewers rely on when choosing what movies to watch (Topal & Ozsoyoglu, 2016).

This analysis aims to investigate which properties of films influence whether they are rated by IMDB as greater than 7 or not. In particular, this analysis examines how factors such as the film's release year, duration, production budget, number of viewer votes and genre are associated with the probability of receiving a high rating.

To investigate this problem, our group first conducted exploratory data analysis to identify potential associations between the variables, and then used a Generalised Linear Model(GLM). Since the response variable indicates whether a film is rated above 7, our group used a binary logistic regression model for the analysis.

# 2 Data Description

The dataset used in this project is dataset06.csv. The dataset contains information of films from the IMDB database and each observation in the dataset represents a single film.

The dataset includes several variables describing film characteristics, including the year of release, film duration, production budget, number of viewer votes and film genre. In addition, the dataset includes the IMDb rating for each film.

Using these variables, our group can investigate the possible correlation between various film attributes and the probability of receiving a rating of 7 or higher.

•film.id – The unique identifier for the film

•year – Year of release of the film in cinemas

•length – Duration (in minutes)

•budget – Budget for the films production (in \$1000000s)

•votes – Number of positive votes received by viewers

•genre – Genre of the film

•rating – IMDB rating from 0-10

```{r}
data <- read.csv("dataset06.csv")

head(data)
str(data)
summary(data)
```

# 3 Data Cleaning

Before conducting the analysis, several data cleaning steps were performed to prepare the dataset for modelling.

First, observations containing missing values were removed using `na.omit()` to ensure that incomplete records did not affect the analysis.

Next, a binary response variable `high_rating` was created from the original rating variable. Films with ratings greater than 7 were coded as 1, while films with ratings of 7 or below were coded as 0.

Finally, the number of votes was log-transformed to reduce skewness and improve the interpretability of the variable.

The cleaned dataset was saved as `Group_6_Data_1.csv`, which is used in the subsequent analysis and modelling steps.

```{r}
df <- read.csv("dataset06.csv")

# remove missing values
df <- na.omit(df)

# create binary response variable
df$high_rating <- ifelse(df$rating > 7,1,0)

# log transform votes
df$log_votes <- log(df$votes)

table(df$genre)
table(df$high_rating)
```

# 4 Exploratory Data Analysis

To explore the relationships between film characteristics and the likelihood of receiving a rating greater than 7, several exploratory visualisations were produced using the cleaned dataset.

```{r}
#| label: fig-genre
#| fig-cap: "Proportion of films rated above 7 across different genres."
#| fig-width: 5
#| fig-height: 4
#| out-width: "70%"
library(ggplot2)
# create bar plot showing proportion of high-rated films by genre
ggplot(df, aes(x = genre, fill = factor(high_rating))) + 
  geom_bar(position = "fill") + 
  labs(
    title = "Proportion of High Rating by Genre",
    x = "Genre",
    y = "Proportion",
    fill = "Rating > 7"
  ) +
  theme_minimal()
```

As shown in @fig-genre, the proportion of films with ratings above and below 7 across different genres.

Short films and documentaries have the highest proportions of high-rated films, followed by animation and comedy. In contrast, drama and romance show very low proportions of films rated above 7. Action films also show a relatively low proportion of high-rated films.

This suggests that genre may be associated with the likelihood of a film receiving a high IMDb rating.

```{r}
#| label: fig-votes
#| fig-cap: "Distribution of log-transformed votes by rating group."
#| fig-width: 5
#| fig-height: 4
#| out-width: "70%"
library(ggplot2)
# create histogram to compare distribution of log-transformed votes
ggplot(df, aes(x = log_votes, fill = factor(high_rating))) + 
  geom_histogram(bins = 30, alpha = 0.7) + 
  labs(
    title = "Log(Votes) by Rating Group",
    x = "log(Votes)",
    fill = "Rating > 7"
  ) +
  theme_minimal()

```

@fig-votes shows the distribution of the log-transformed number of votes for films with rating over and below 7.

For both groups, most films fall within the lower ranges of log(votes), indicating that regardless of their ratings, only a small number of films received a very high number of votes, most films received a moderate number of votes.

Although the red bars (ratings ≤ 7) are consistently higher than the blue bars (ratings \> 7), this does not mean that lower-rated films are more likely to receive more votes.

First, films rated above 7 appear to account for a smaller share of the overall dataset. So the taller red bars are results of the difference in total group sizes.

Second, the distributions of the two groups overlap substantially, both extending from approximately log(Votes) ≈ 2 to over 10. This suggests that the number of votes does not strongly distinguish between high-rated and low-rated films.

```{r}
#| label: fig-length
#| fig-cap: "Distribution of film length across rating groups."
#| fig-width: 5
#| fig-height: 4
#| out-width: "70%"
library(ggplot2)
# create boxplot to compare film length across rating groups
ggplot(df, aes(x = factor(high_rating), y = length, fill = factor(high_rating))) +
  geom_boxplot() +
  scale_x_discrete(labels = c("0" = "Rating ≤ 7", "1" = "Rating > 7")) +
  labs(
    title = "Film Length by Rating Group",
    x = "Rating Group",
    y = "Length (mins)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

As shown in Figure @fig-length, the distribution of film length differs slightly between films with ratings above 7 and those with ratings of 7 or below.

The median film length of lower-rated films appears slightly higher than the highly rated films.

In addition, the distribution of film length among highly rated films appears more variable, with a wider spread of values.

Several longer films are also observed among the lower-rated group as outliers.

While the exploratory analysis focused on selected key variables such as genre, film length, and log-transformed votes, other variables including year and budget were not visualised due to space constraints. However, these variables were still included in the regression model as they are theoretically relevant and may contribute to explaining variation in film ratings.

Overall, the boxplot suggests that film length may be associated with rating, although the relationship does not appear to be very strong.

# 5 Model (GLM)

To examine how film properties influence the probability of receiving a rating above 7, a logistic regression model was fitted. Since the response variable high_rating is binary, a GLM with a binomial family and logit link was used.

## 5.1 Model specification

Since the response variable `high_rating` is binary, a logistic regression model is used. Let $p_i$ denote the probability that film $i$ receives a rating above 7.

The model assumes

$$
Y_i \sim \text{Binomial}(1, p_i)
$$

and uses the logit link function as Model 1.

Model 1 includes the predictors: year, length, log_votes, and budget.

$$
\log\left(\frac{p_i}{1-p_i}\right)
=
\beta_0
+
\beta_1 year_i
+
\beta_2 length_i
+
\beta_3 log\_votes_i
+
\beta_4 budget_i
$$

Model 2 additionally includes the categorical variable `genre`.
$$
\log\left(\frac{p_i}{1-p_i}\right)
=
\beta_0
+
\beta_1 \text{year}_i
+
\beta_2 \text{length}_i
+
\beta_3 \text{log\_votes}_i
+
\beta_4 \text{budget}_i
+
\sum_{j} \gamma_j \text{genre}_{ij}
$$

where $\text{genre}_{ij}$ represents dummy variables for each genre category, and one category (Action) is used as the reference group.

In this model, the inclusion of genre allows for differences in the probability of receiving a high rating across film types. Each genre coefficient represents the change in log-odds relative to the reference category, holding other variables constant.

## 5.2 Model fitting

The two models were fitted in R as follows.

```{r}
model1 <- glm(
  high_rating ~ year + length + log_votes + budget,
  family = binomial(link = "logit"),
  data = df
)

model2 <- glm(
  high_rating ~ year + length + log_votes + budget + genre,
  family = binomial(link = "logit"),
  data = df
)
```

## 5.3 Model comparison

The two models were compared using the Akaike Information Criterion (AIC).

```{r}
#| label: tbl-aic
#| tbl-cap: "Comparison of AIC values for Model 1 and Model 2."
#| echo: false

# compute AIC values for both models
aic_table <- AIC(model1, model2)

# display results as a formatted table
knitr::kable(aic_table)
```

As shown in @tbl-aic, Model 2 provides a better fit.

Model 2 has a considerably lower AIC than Model 1, suggesting that the inclusion of `genre` improves the fit of the model.

## 5.4 Selected model

To further assess the models, the deviance values were also examined. The null deviance for Model 2 was 2327.27, while the residual deviance was reduced to 947.91. This substantial reduction indicates that the predictors included in the model explain a considerable amount of variation in whether a film receives a rating above 7.

In addition, the large improvement in AIC from Model 1 to Model 2 suggests that `genre` contributes important explanatory information beyond the continuous predictors alone. Therefore, genre should be retained in the final model.

## 5.5 Selected model

Based on the AIC comparison and the reduction in deviance, Model 2 was selected as the final model for interpretation. This model provides a better representation of the factors associated with high IMDb ratings and is therefore used in the Results section.

# 6 Finding

```{r}
#| label: tbl-model2
#| tbl-cap: "Coefficient estimates for the final logistic regression model."
#| echo: false

# extract coefficient estimates from model summary
coef_table <- summary(model2)$coefficients

# display coefficient table
knitr::kable(coef_table)
```

According to @tbl-model2, several predictors are statistically significant.

Model 2, which includes `genre` as an additional predictor, provides a substantially better fit than Model 1, with a much lower AIC (969.91 compared with 1594.81). This indicates that genre plays an important role in explaining whether a film receives an IMDb rating above 7.

From the fitted logistic regression model, several predictors are statistically significant.

The coefficient for `year` is positive and statistically significant (p = 0.00879), suggesting that more recently released films are slightly more likely to receive a rating above 7, holding other variables constant.

The coefficient for `length` is negative and highly statistically significant (p \< 0.001), indicating that longer films are less likely to receive a high rating.

The coefficient for `budget` is positive and highly statistically significant (p \< 0.001), implying that films with larger production budgets are more likely to achieve ratings above 7.

In contrast, `log_votes` is not statistically significant (p = 0.8506), indicating that after controlling for the other variables, the number of votes does not meaningfully affect the probability of a high rating.

For the categorical variable `genre`, Action is the reference category. Relative to Action films, Comedy, Documentary, and Short films have significantly higher odds of receiving a rating above 7 (all p \< 0.01), while Drama films have significantly lower odds (p \< 0.001). Animation and Romance are not significantly different from Action films at the 5% significance level.

Overall, the results show that `genre`, `budget`, `length`, and `year` are important determinants of high IMDb ratings, while `log_votes` is not significant in the presence of the other variables.

# 7 Discussion

This study investigated which film characteristics influence whether a film receives an IMDb rating above 7 using a logistic regression model.

The results demonstrate that `genre` is one of the most important factors. Films such as Documentary, Comedy, and Short are significantly more likely to receive high ratings compared to Action films, while Drama films are less likely to achieve high ratings. This suggests that audience preferences and evaluation standards vary across genres.

The analysis also shows that `budget` has a strong positive effect, indicating that higher-budget films are more likely to receive high ratings. This may be because larger budgets allow for better production quality, stronger marketing, and wider distribution.

In addition, `length` has a negative effect, suggesting that longer films are less likely to receive ratings above 7. This may indicate that shorter films are more engaging or easier for audiences to evaluate positively.

Although the number of votes might be expected to reflect popularity, `log_votes` is not statistically significant in the model. This suggests that the number of votes alone does not determine whether a film receives a high rating once other factors are taken into account.

The effect of `year` is positive but relatively small, indicating that more recent films are slightly more likely to receive higher ratings.

## Limitations

There are several limitations to this analysis. First, the dataset does not include potentially important variables such as director, cast, awards, or marketing, which may influence film ratings. Second, the model assumes linear relationships between predictors and the log-odds of a high rating, which may oversimplify real-world relationships. Finally, this analysis is based on observational data and cannot establish causal relationships.

## Conclusion

In conclusion, this analysis shows that `genre`, `budget`, `length`, and `year` significantly influence whether a film receives an IMDb rating above 7. Among these, genre appears to be the most important factor. These findings provide useful insights for understanding the characteristics associated with highly rated films.

# 8 References

-   K. Topal and G. Ozsoyoglu, "Movie review analysis: Emotion analysis of IMDb movie reviews," 2016 IEEE/ACM International Conference on Advances in Social Networks Analysis and Mining (ASONAM), San Francisco, CA, USA, 2016, pp. 1170-1176, doi: 10.1109/ASONAM.2016.7752387.

#
